---
title: "Clean the sample metadata"
author: Michael McLaren
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
---


```{r setup, include=FALSE}
knitr::opts_knit$set(progress = TRUE, verbose = TRUE)
# Global chunk options
knitr::opts_chunk$set(
    include = TRUE, echo = TRUE,
    warning = TRUE, message = FALSE, 
    fig.width = 6, fig.asp = 3/4
)
```

```{r load_packages}
library(tidyverse)
library(here)
```

```{r}
sam <- readxl::read_excel(here("data", "Metadata_2.13.2019.xlsx"),
    na = c("", "NA", "Unknown"))
glimpse(sam)
```
Some quick variable-name reformatting to make our job easier:
```{r}
sam <- sam %>%
    # replace the problematic `SN #` var
    rename("Sample_number" = `SN #`) %>%
    # standardize case
    rename_all(~ifelse(str_detect(., " "), str_to_sentence(.), .)) %>%
    # replace spaces with underscores
    rename_all(~str_replace_all(., "\\.? ", "_"))
glimpse(sam)
```
The last column lacks a name and is mostly empty; it only records a note about
one sample that has no metadata,
```{r}
sam %>%
    filter(!is.na(...18)) %>%
    unlist
```
There are two samples with no metadata and which were not analyzed.
```{r}
sam %>%
    filter(is.na(Species) | is.na(C_diff_lab_positive)) %>%
    select(Sample_number:Species, C_diff_lab_positive, everything())
```
We will drop these samples now, and the unnecessary column
```{r}
sam <- sam %>%
    filter(!is.na(Species)) %>%
    select(-...18)
```

## Check and clean individual variables

#### Sample identifiers

```{r}
sam <- sam %>%
    mutate_at("Sample_number", as.integer) %>%
    # It is safer to replace the dashes with underscores in the sample names
    # for some analyses in R
    mutate(Sample = str_replace(Sample, "-", "_")) %>%
    # IDs submitted to sequencing center (??)
    rename(Submitted_ID = ID) %>%
    # IDs in the community sequence data (fastq) files
    mutate(Sequence_ID = str_replace(Submitted_ID, "_", "-"))
```

#### Year sequenced

```{r}
sam %>%
    group_by(Microbiome) %>%
    count
sam <- sam %>%
    rename(Year_sequenced = Microbiome) %>%
    mutate(
        Year_sequenced = factor(Year_sequenced, c("2017", "2018")),
        Year_sequenced = fct_explicit_na(Year_sequenced, "Not sequenced")
        )
sam %>%
    group_by(Year_sequenced) %>%
    count
```

Which samples were not sequenced?
```{r}
sam %>%
    filter(Year_sequenced == "Not sequenced") %>%
    select(Sample, Submitted_ID, Species, C_diff_lab_positive)
```

Addendum by MM: The samples were sequenced in two different submissions to the
Michigan Host Microbiome Initiative (HMI), and this field designates which
submission the sample was in. Though I called this field `Year_sequenced`, the
"2018" samples were actually sequenced in late 2017, and so this field really
denotes when the data was first analyzed by NB and RT. Also, some of the
samples marked "Not sequenced" were submitted for sequencing, but did not yield
useable data because of failed library prep or sequencing or because of fears
of contamination or mislabeling.

#### Host species

Change `Species` to `Host_species` to avoid conflict with the `Species`
variable denoting the microbial ASV species when doing microbiome analysis
```{r}
sam <- sam %>%
    rename(Host_species = Species)
sam %>%
    group_by(Host_species) %>%
    count
```
Note that 1 Alpaca and 1 Avian sample are included but not used in our
analyses.

#### Sex

Currently, the `Sex` variable combines sex (M or F) with neuter status (empty
string "" for not neutered, and "S" or "C" for spayed or castrated)
```{r}
sam %>%
    group_by(Sex) %>%
    count
```
We will keep this original classification in a variable `SexCS` but also create
separate Sex and Neutered fields,
```{r}
sam <- sam %>%
    rename(SexCS = Sex) %>%
    mutate(
        Sex = str_sub(SexCS, 1, 1) %>% factor,
        Neutered = fct_collapse(str_sub(SexCS, 2, 2),
            "No" = "", "Yes" = "C", "Yes" = "S") %>%
            fct_relevel("No")
    ) %>%
    mutate_at(vars(SexCS, Sex, Neutered), fct_explicit_na, "Not known")
sam %>%
    select(SexCS, Sex, Neutered) %>%
    distinct %>%
    arrange(Sex, Neutered)
```

Note about binary variables: It will be useful to have all binary variables in
a compatible format that matches the order of the No/Negative type with the
Yes/Positive type. We could do this by converting them to factors or converting
to logical or 0/1 variables. For now I'll go with logical variables.

#### GI status

Give factor levels consistent with the main tables
```{r}
sam %>% group_by(GI) %>% count
sam <- sam %>%
    rename(GI_status = GI) %>%
    mutate(GI_status = fct_recode(GI_status,
            Healthy = "No",
            `GI condition` = "Yes"
            ) %>%
            fct_explicit_na("Not known")
        )
sam %>% group_by(GI_status) %>% count
```

#### Antibiotics

There are three antibiotics fields; let's give them a uniform naming scheme,
and give factor levels for antibiotics usage consistent with the main table.
```{r}
sam <- sam %>%
    rename(
        Ab_status = Ab,
        Ab_type = Antibiotics,
        Ab_group = Antibiotic_group
    )
sam %>%
    group_by(Ab_status) %>%
    count
sam <- sam %>%
    mutate(Ab_status = factor(Ab_status, c("Non-treated", "Treated")) %>%
            fct_explicit_na("Not known"))
sam %>%
    group_by(Ab_status) %>%
    count
```

Note: Antibiotics type and group are inconsistently reported, but we won't be
making use of them in our microbiome analysis and so will be leaving as is.

#### C. difficile

There are three C difficile variables; let's give them a uniform naming scheme.
```{r}
sam <- sam %>%
    rename(
        CD_lab = C_diff_lab_positive,
        CD_toxigenic = Toxigenic,
        CD_ribotype = Ribotype,
    )
```
and convert the CD lab assay results to a factor,
```{r}
sam %>% group_by(CD_lab) %>% count
sam <- sam %>%
    mutate(CD_lab = factor(CD_lab, c("Negative", "Positive")))
sam %>% group_by(CD_lab) %>% count
```
CD toxigenic status factors like in the main table,
```{r}
sam %>% group_by(CD_toxigenic) %>% count
sam <- sam %>%
    mutate(CD_toxigenic = fct_recode(CD_toxigenic,
        `Non-toxigenic` = "No",
        `Toxigenic` = "Yes")
    ) 
sam %>% group_by(CD_toxigenic) %>% count %>% arrange(CD_toxigenic)
```

Inspect the ribotypes:
```{r}
sam %>% group_by(CD_lab, CD_ribotype) %>% count
sam %>% filter(is.na(CD_ribotype)) %>% select(Sample_number, Sample,
    Host_species)
```
5 samples for which Cdiff was recovered are `NA` for ribotype. Three of these
samples (CS77-CN47, CS98-CN57, CS185-CN102) were sequenced but not in the
database, and two of these samples (CS104-EQ28, CS186-CN103) were not sequenced
successfully.
```{r}
sam <- sam %>%
    mutate(CD_ribotype = case_when(
            Sample %in% c("CS77_CN47", "CS98_CN57", "CS185_CN102") ~ "Not in database",
            Sample %in% c("CS104_EQ28", "CS186_CN103") ~ "Not sequenced",
            TRUE ~ CD_ribotype
            ))
sam %>% group_by(CD_lab, CD_ribotype) %>% count
```

#### Age

Age is listed as days, months, or years,
```{r}
sam$Age %>% unique %>% unlist %>% sort
sam <- sam %>%
    # Fix typos
    mutate(Age = str_replace(Age, "yera", "year"))
```
Let's convert age to a numeric variable in years.
```{r}
age <- sam %>%
    select(Age) %>%
    distinct %>%
    separate(Age, c("Number", "Unit"), sep = " ", 
        remove = FALSE) %>%
    mutate(
        Number = as.numeric(Number),
        Age_years = case_when(
            str_detect(Unit, "day") ~ Number / 365,
            str_detect(Unit, "month") ~ Number / 12,
            str_detect(Unit, "year") ~ Number,
            ))
age %>% arrange(Age_years) %>% print(n=Inf)
# Note that we kept a row for Age = NA for the join in the next step
# Join w/ the sample metadata, and set Age to age in years
sam <- left_join(sam, age %>% select(Age, Age_years), by = "Age") %>%
    rename(Age_reported = Age, Age = Age_years)
```

Let's check the manually input age grouping for canines against the Age
(canines only):
```{r}
sam %>%
    group_by(Host_species, Age_group) %>%
    count
sam %>%
    group_by(Age_group) %>%
    summarize(min(Age), max(Age))
```
Set the age group as a factor,
```{r}
sam <- sam %>%
    mutate(
        Age_group = ifelse(Host_species == "Canine", Age_group, 
            "Not applicable"),
        Age_group = factor(Age_group, 
            c("Not applicable", "<1", "1-4", "5-9", "≥10")) %>%
            fct_explicit_na("Not known")
    )
sam %>%
    group_by(Host_species, Age_group) %>%
    count %>%
    arrange(Age_group)
```

## Sort and save

```{r, eval = F}
sam <- sam %>%
    select(Sample, Host_species, Breed, starts_with("Age"), 
    starts_with("Sex"), Neutered, 
    starts_with("Ab"), starts_with("GI"), starts_with("CD"),
    Relevant_diagnosis_or_reason_for_visit,
    Sample_number, Sequence_ID, Year_sequenced)
write_csv(sam, here("results", "metadata-clean.csv"))
saveRDS(sam, here("results", "metadata-clean.Rds"))
```
