---
title: "Compositional PCA and differential-abundance analysis"
author: Michael McLaren
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
---

# Setup

```{r setup, include=FALSE}
knitr::opts_knit$set(progress = TRUE, verbose = TRUE)
# Global chunk options
knitr::opts_chunk$set(
    cache = TRUE, autodep = TRUE,
    include = TRUE, echo = TRUE,
    warning = TRUE, message = FALSE, 
    fig.width = 6, fig.asp = 3/4
)
```

```{r load_packages}
library(here)
library(phyloseq)
library(tidyverse)
library(ggbeeswarm)
library(ggthemes)
library(factoextra)
library(cowplot)
library(ALDEx2)
# Custom functions and plot setup
source(here("analysis", "functions-and-ggplot-setup.R"))
```

Set different ggplot theme for this file,
```{r}
theme_set(theme_minimal() + theme(panel.grid = element_blank()))
color.axis <- "darkgrey"
```

```{r, load_data}
ps <- readRDS(here("results", "ps.Rds"))
# Make variable CD to be our indicator of Cdiff presence for this analysis
sample_data(ps)$CD <- sample_data(ps)$CD_either
# Shorten the sample names
sample_names(ps) <- sample_names(ps) %>%
    str_match("(.+)_") %>% 
    .[,2]
sam <- sample_data(ps) %>% as_tibble
tax <- tax_table(ps) %>% as_tibble
```

Restrict our analysis to canine samples that appear in at least 5 samples and
at least 100 reads.
```{r}
filt_fun <- function (x, min_reads = 100, min_samples = 5) {
    (sum(x) > min_reads) & (sum(x > 0) > min_samples)
}
ps.canine <- subset_samples(ps, Host_species == "Canine") %>%
    filter_taxa(filt_fun, prune = TRUE)
```

# CD associations in canines

## Presence/absence testing

Viewing the CLR abundances of the top hits (see below) suggests that the top
associations detected by ALDEx2 are driven mainly by differential prevalence
(presence/absence), rather than a consistent log-fold change in abundance
(which is what ALDEx2 and most other differential abundance methods are
designed to measure). We can test for differential prevalence directly with a
Fisher's Exact Test, which we'll compare to the ALDEx2 associations below.
```{r}
# Data frame with the prevalence of each taxon (ASV) in CD+ and CD- samples
pa <- ps.canine %>% as_tibble %>%
    group_by(Taxon, CD) %>% 
    summarize( 
        Absent = sum(Abundance == 0), 
        Present = sum(Abundance > 0),
        Prevalence = mean(Abundance > 0)
    ) %>%
    ungroup
# For each ASV, Fisher's exact test of association with CD
pa0 <- pa %>%
    select(-Prevalence) %>%
    group_by(Taxon) %>%
    nest %>%
    mutate(test = map(data, ~fisher.test(.[,2:3]))) %>%
    unnest(map(test, broom::tidy)) %>%
    select(-data, -test)
# Top 10 positive and negative associations
pa0 %>%
    group_by(estimate > 1) %>%
    top_n(10, -p.value) %>%
    arrange(estimate < 1, p.value) %>%
    select(-method,-alternative) %>%
    knitr::kable()
```

## Aldex

```{r x.canine}
set.seed(20190314)
x <- aldex.clr(otu_table(ps.canine) %>% t, 
    sample_data(ps.canine)$CD,
    mc.samples = 128, denom = "all", useMC = TRUE)
x.tt <- aldex.ttest(x, paired.test = FALSE) %>%
    as_tibble(rownames = "Taxon")
x.effect <- aldex.effect(x, useMC = TRUE) %>%
    as_tibble(rownames = "Taxon")
x.all <- left_join(x.tt, x.effect, by = "Taxon")
remove(x.tt, x.effect)
# denom <- getDenom(x) %>%
#     getFeatures(x)[.] %>%
#     names
```

Aldex "MW" plot
```{r, dependson = "x.canine", fig.width = 8}
cutoff = 0.1
aldex.plot(x.all, type = "MW", test = "wilcox", cutoff = cutoff)
```

Examine the top positive and negative associations
```{r, dependson = "x.canine.pd"}
top_hits <- x.all %>%
    left_join(tax, by = "Taxon") %>%
    group_by(sign(effect)) %>%
    top_n(15, abs(effect)) %>%
    ungroup %>%
    left_join(tax) %>%
    select(Taxon, effect, starts_with("w"), Phylum, Genus) %>%
    arrange(-sign(effect), -abs(effect))
top_hits %>%
    knitr::kable(digits = 3)
```

Compare the ALDEx2 associations with differential prevalence for the top ALDEx2
hits,
```{r}
top_hits %>%
    left_join(pa0 %>% select(Taxon, estimate, p.value), by = "Taxon") %>%
    select(-Phylum) %>%
    knitr::kable(digits = 3)
```
The top differentially abundant taxa according to ALDEx2 mainly have p-valuse <
0.05 by the Fisher test, and in the expected direction. This observation is
consistent with the idea that ALDEx2 is mainly picking up presence/absence
patterns, which we explore visually below.

```{r x.canine.pd, dependson = "x.canine"}
# Pick the top 4 positive and negative associations, plus ASV62
taxa <- x.all %>%
    group_by(sign(effect)) %>%
    top_n(4, abs(effect)) %>%
    arrange(-sign(effect), -abs(effect)) %>%
    .$Taxon %>%
    c("ASV62")
clr.post <- getMonteCarloInstances(x) %>%
    map(t) %>%
    map_dfr(~as_tibble(.) %>% mutate(idx = row_number()), .id = "Sample") %>%
    select(Sample, idx, taxa) %>%
    gather("Taxon", "CLR", taxa) %>%
    left_join(sam, by = "Sample") %>%
    left_join(tax, by = "Taxon")
clr.post <- clr.post %>%
    mutate(Taxon = factor(Taxon, taxa))
clr.means <- clr.post %>%
    group_by_at(vars(-idx, -CLR)) %>%
    summarize(CLR = mean(CLR)) %>%
    ungroup
```

Labels for the top taxa by genus and species; use the NCBI organism for C.
hiranonis; and only use the first identified species for Silva species id in
cases where there are multiple matches.
```{r}
taxlabs <- tax %>%
    filter(Taxon %in% taxa) %>%
    mutate(
        Species = str_extract(Species, "[^/]+"), # Grab just the first species
        Label = case_when(
            !is.na(NCBI_species) ~ NCBI_species,
            TRUE ~ paste(Genus, Species)
        )
    ) %>%
    mutate(Taxon = factor(Taxon, taxa),
        Label = paste(Taxon, Label, sep = "\n"),
        # Put Genus + species on different lines
        Label = str_replace(Label, " ", "\n"),
        ) %>%
    select(Taxon, Label) %>%
    deframe
taxlabs
```

In figure legend, can note that hiranonis is classified as Peptoclostridium
(genus) and that ASV112 matches three Sutterella species
(massiliensis/stercoricanis/wadsworthensis).

```{r, dependson = "x.canine.pd", fig.width = 8, fig.asp = 4/4}
pa1 <- pa %>%
    filter(Taxon %in% taxa) %>%
    mutate(Prevalence = round(Prevalence, 2), Taxon = factor(Taxon, taxa))
p.violin <- clr.post %>%
    filter(idx <= 32) %>%
    ggplot(aes(CD, CLR, color = CD)) +
    # geom_hline(yintercept = 0.4, color = color.axis) +
    # geom_violin(draw_quantiles = c(0.5), bw = 0.5) +
    geom_violin(bw = 0.5) +
    geom_quasirandom(data = clr.means, size = 0.8) +
    scale_y_continuous(limits = c(-6, 17), breaks = c(-5, 0, 5, 10),
        oob = scales::squish) +
    facet_grid(Taxon ~ ., labeller = labeller(Taxon = taxlabs)) +
    coord_flip() +
    geom_text(data = pa1, aes(label = Prevalence), y = 16.7) +
    theme(
        # panel.grid.major.y = element_line(color = color.axis),
        legend.position = "top",
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.y = element_text(angle = 0),
        panel.spacing = unit(1, "lines")
        ) +
    labs(y = "CLR relative abundance")
p.violin
```

It is difficult to tell which points correspond to samples in which the taxon
has a read count of 0, since the mean CLR value for a taxon with a read count
of 0 depends on the read depth and sample composition. Let's see what the
largest CLR value is for such a taxon.
```{r}
tb <- clr.means %>% 
    select(Sample, Taxon, CD, CLR) %>%
    left_join(otu_table(ps.canine) %>% as_tibble(tidy = TRUE), 
        by = c("Sample", "Taxon"))
tb %>%
    filter(Abundance == 0) %>%
    arrange(-CLR)
```
In one case, the mean CLR is positive (0.276), but in the rest of the cases it
is negative.

```{r, fig.width = 8}
tb %>%
    filter(Abundance <= 10) %>%
    ggplot(aes(CD, CLR, color = Abundance > 0)) +
    facet_grid(Taxon ~ .) +
    geom_hline(yintercept = 0) +
    geom_hline(yintercept = 0.3) +
    geom_quasirandom() +
    coord_flip() 
```


# PCA

Weighted pca to give even weight to CD+/- groups

note, weighted pca looks almost identical to the unweighted pca, though the
taxa (loadings?) are somewhat larger

```{r pca.canine, dependson = "x.canine"}
# Weight by 1 / # of samples in group
ws <- sample_data(ps.canine) %>%
    as_tibble %>%
    group_by(CD) %>%
    mutate(Weight = 1 / n()) %>%
    ungroup %>%
    select(Sample, Weight) %>%
    deframe
# Weighted pca
pcx.ws <- otu_table(ps.canine) %>%
    transform_sample_counts(function (x) clr(x + 0.5)) %>%
    ade4::dudi.pca(row.w = ws, center = TRUE, scale = FALSE, scannf = FALSE,
    nf = 5)
# Unweighted PCA
pcx <- otu_table(ps.canine) %>%
    transform_sample_counts(function (x) clr(x + 0.5)) %>%
    ade4::dudi.pca(center = TRUE, scale = FALSE, scannf = FALSE,
    nf = 5)
# Percent variance explained
perc_var <- pcx$eig %>% 
    {. / sum(.)} %>%
    head(n=5) %>%
    round(., 2) %>%
    {paste0("[", 100 * . , "%]")}
perc_var
# Samples plot
pctb.sam <- fviz_pca_ind(pcx) %>%
    .$data %>% 
    as_tibble() %>%
    rename(Sample = name, PC1 = x, PC2 = y) %>%
    left_join(sam, by = "Sample")
p1 <- pctb.sam %>%
    ggplot(aes(PC1, PC2, color = CD)) +
    geom_hline(yintercept = 0, color = color.axis) +
    geom_vline(xintercept = 0, color = color.axis) +
    geom_point() +
    theme(legend.position = "bottom") +
    scale_x_continuous(limits = range(pctb.sam$PC1)) +
    scale_y_continuous(limits = range(pctb.sam$PC2)) +
    labs(title = "PCA ordination: Samples", 
        x = paste("PC1", perc_var[1]), y = paste("PC2", perc_var[2]))
# Taxa plot
pctb.taxa <- fviz_pca_var(pcx) %>%
    .$data %>% 
    as_tibble() %>%
    rename(Taxon = name, PC1 = x, PC2 = y) %>%
    left_join(tax, by = "Taxon") %>%
    left_join(x.all, by = "Taxon")
p2 <- pctb.taxa %>%
    filter(Taxon %in% top_hits$Taxon) %>%
    ggplot(aes(PC1, PC2, label = Taxon, color = as.factor(sign(effect)))) +
    geom_hline(yintercept = 0, color = color.axis) +
    geom_vline(xintercept = 0, color = color.axis) +
    geom_point(aes(size = abs(effect))) +
    ggrepel::geom_text_repel(data = pctb.taxa %>% filter(Taxon %in% taxa)) +
    theme(legend.position = "bottom") +
    scale_x_continuous(limits = range(pctb.sam$PC1)/6) +
    scale_y_continuous(limits = range(pctb.sam$PC2)/6) +
    scale_size(breaks = c(0.3, 0.6, 0.9)) +
    guides(color = guide_legend(override.aes = list(label = ""))) +
    labs(title = "PCA ordination: ASVs (6X zoom)", color = "sign(effect)",
        x = paste("PC1", perc_var[1]), y = paste("PC2", perc_var[2]))
# pctb.taxa %>% filter(wi.ep < 0.05) %>% .$effect %>% abs %>% range
```

```{r, dependson = "pca.canine", fig.width = 10, fig.asp = 1/2}
cowplot::plot_grid(
    p1,
    p2,
    align = "h"
)
```

### Possible main text figure

```{r, fig.width = 9, fig.asp = 0.8}
pl <- plot_grid(
    p1 + theme(legend.position = "bottom") +
        labs(color = "C. difficile presence"),
    p2 + theme(legend.position = "bottom") + 
        guides(color = FALSE, size = guide_legend()),
    align = "v", ncol = 1,
    labels = c("A", "B")
)
# plot_grid(pl, p.violin, axis = "tb", align = "h")
p <- plot_grid(pl, NULL, 
    p.violin + 
        theme(legend.position = "bottom") + 
        guides(color = guide_legend(override.aes = list(label = ""))) +
        labs(title = "       Top differentially-abundant ASVs", 
            x = "Relative abundance (CLR)",
            color = "C. difficile presence"),
    nrow = 1, rel_widths = c(1, 0.0, 1), 
    labels = c("", "", "C"), label_x = -0.00
    )
    # draw_text("Prevalence", x = 0.895, y = 0.98, hjust = 1, vjust = 1, 
    #     size = 12)
ggdraw(p) + 
    draw_text("Prevalence", x = 0.893, y = 0.072, hjust = 1, vjust = 0,
    size = 11)
ggsave(here("figures", "community-cd-association.pdf"), 
    width = 7, height = 5.5, units = "in", scale = 5/3,
    useDingbats = FALSE)
```
