---
title: "Logistic regression of C. difficile ~ C. hiranonis"
author: Michael McLaren
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
---

## Setup

```{r setup, include=FALSE}
knitr::opts_knit$set(progress = TRUE, verbose = TRUE)
# Global chunk options
knitr::opts_chunk$set(
    cache = TRUE, autodep = TRUE,
    include = TRUE, echo = TRUE,
    warning = TRUE, message = FALSE, 
    fig.width = 6, fig.asp = 3/4
)
```

```{r load_packages}
library(here)
library(phyloseq)
library(tidyverse)
library(ggbeeswarm)
library(ggthemes)
library(ggstance)
library(cowplot)
library(brms)
library(tidybayes)
# Custom functions and plot setup
source(here("analysis", "functions-and-ggplot-setup.R"))
```

```{r, load_data}
ps <- readRDS(here("results", "ps.Rds")) %>%
    subset_samples(Host_species %in% c("Canine", "Feline", "Equine", 
            "Ovine")) %>% 
    filter_taxa(function (x) sum(x) > 0, prune = TRUE)
# Make variable CD to be our indicator of Cdiff presence for this analysis
sample_data(ps)$CD <- sample_data(ps)$CD_either
# Shorten the sample names
sample_names(ps) <- sample_names(ps) %>%
    str_match("(.+)_") %>% 
    .[,2]
sam <- sample_data(ps) %>% as_tibble
```

Get a data frame formatted for the regressions
```{r}
tb <- ps %>%
    subset_samples(Host_species == "Canine") %>%
    prune_taxa(c("ASV4", "ASV62"), .) %>%
    as_tibble(tax = FALSE) %>%
    spread(Taxon, Abundance) %>%
    mutate(
        CH = factor((ASV4 > 0) + (ASV62 > 0)),
        ASV4_present = ASV4 > 0,
        ASV62_present = ASV62 > 0
    )
```
The variable `CH` indicates whether 0, 1, or 2 of the two ASVs are present in
the sample; thus `CH==1` and `CH==2` correspond to the two putative C. hiranonis
strains. Since ASV4 is always present if any C. hiranonis ASV is, we can use
ASV4 presence as an indicator of C. hiranonis presence.

Set the seed,
```{r}
set.seed(42)
```

## Failure of standard logistic regression due to perfect separation

Since all canines with ASV62 are CD-, standard logistic regression with the
`glm` function can't be used to estimate the effect of ASV62 or CH2,
```{r}
m1 <- glm(data = tb, family = binomial, CD ~ CH)
m1
confint(m1)
```

## Bayesian logistic regression

We can solve this problem by using Bayesian logistic regression with
weakly-informative priors to bound the effect sizes.

### Picking priors

We'll use a t distribution with 7 degrees of freedom with the spread set so
that the 0.01 and 0.99 quantiles on the baseline Cdiff prevalence are roughly
0.01 and 0.99 (e.g., less than 0.01 probability of a prevalance < 0.01),
```{r}
df_t <- 7
sigma_t <- 1.53
inv_logit_scaled(qt(c(0.01, 0.1, 0.9, 0.99), df_t) * sigma_t)
```
and that the odds ratios are bounded so that the 0.01 and 0.99 quantiles
correspond to odds ratios of 1/200 and 200,
```{r}
df_t <- 7
sigma_t <- 1.77
exp(qt(c(0.01, 0.1, 0.9, 0.99), 7) * sigma_t)
```

```{r}
base_prior <- c(
    prior(student_t(7, 0, 1.53), class = Intercept),
    prior(student_t(7, 0, 1.77), class = b)
)
```

### C. hiranonis-only model

```{r}
b1 <- brm(data = tb, family = bernoulli, CD ~ CH, 
    prior = base_prior)
b1
```
Here, CH1 and CH2 denote the effects of the putative different strains of C.
hiranonis. 

Note, we expect the estimate for CH2 is sensitive to the width of the prior due
to the lack of any CD+ canines with ASV62.

Baseline prevalence:
```{r}
b1 %>% 
    fixef(pars = "Intercept", probs = c(0.05, 0.1, 0.9, 0.95)) %>% 
    inv_logit_scaled
```
Coefficients in log-odds space:
```{r}
b1 %>% 
    fixef(probs = c(0.05, 0.1, 0.9, 0.95))
```
Odds-ratios:
```{r}
b1 %>% 
    fixef(probs = c(0.05, 0.1, 0.9, 0.95)) %>% 
    exp
```

### Antibiotics, GI status, and Sex

Note, there are four samples that are missing information about these
variables, and so we need to first remove these samples.
```{r}
tb %>%
    filter_at(vars(Ab_status, GI_status, Sex), any_vars( . == "Not known")) %>%
    select(Sample, Ab_status, GI_status, Sex)
tb2 <- tb %>%
    filter_at(vars(Ab_status, GI_status, Sex), all_vars(. != "Not known")) %>%
    mutate_at(vars(Ab_status, GI_status, Sex), fct_drop)
tb2 %>%
    group_by(Ab_status, GI_status, Sex) %>%
    count
```
Now let's add Antibiotics, GI status, and Sex to the model,
```{r}
b2 <- brm(data = tb2, family = bernoulli,
    CD ~ Ab_status + GI_status + Sex + CH,
    prior = base_prior
)
b2
```

Plot:
```{r}
lvls <- paste0("b_", c("Intercept", "GI_statusGIcondition", "Ab_statusTreated",
        "SexM", "CH1", "CH2"))
labeller <- function(lvl) {
    lvl %>% 
        str_replace("b_", "") %>% 
        str_replace("_", " ")
}
b2 %>%
    gather_draws(`b_.+`, regex = TRUE) %>%
    ggplot(aes(y = factor(.variable, rev(lvls)), x = .value)) +
    geom_vline(xintercept = 0, color = "grey") +
    stat_intervalh(.width = c(.50, .80, .95, .99)) +
    scale_x_continuous(limits = c(-7, 3), oob = scales::squish) +
    scale_y_discrete(labels = labeller) +
    scale_color_brewer() +
    labs(y = "Variable", x = "Value of coefficient",
        title = "Coefficients from logistic regression of CD+",
        color = "Credible\ninterval") +
    guides(colour = guide_legend(reverse = TRUE)) +
    theme_minimal() + 
    theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())
ggsave(here("figures", "logreg-cd.pdf"), 
    width = 5, height = 3, units = "in")
```

Note, the lower end of the credible interval for the `CH2` coefficient is
clipped at `-7`.

### Age

Now we'll add age to the model. As before, we'll drop the samples without age
information,
```{r}
tb2 %>%
    group_by(is.na(Age)) %>%
    count
tb3 <- tb2 %>%
    filter(!is.na(Age))
```
Note, 26% of the samples have been dropped due to missing ages.

Recall that there appeared to be a non-monotonic effect of age. We can try to
account for this effect with a spline, First, let's test the age-only model.
We'll use the default prior on the spline standard deviation.
```{r}
b.age <- brm(data = tb3, family = bernoulli,
    CD ~ s(Age),
    prior = base_prior,
    control = list(adapt_delta = 0.95)
)
```

Compare the predicted prevalence to the distribution of observed cases:
```{r}
f <- tibble(Age = seq(0, 16, by = 0.1)) %>%
    bind_cols(., fitted(b.age, ., probs = c(0.1, 0.9)) %>% as_tibble)
f %>%
    ggplot(aes(Age, Estimate)) +
    geom_ribbon(aes(ymin = Q10, ymax = Q90), 
        linetype = 0, alpha = 0.2) +
    geom_line() +
    geom_quasirandom(data = tb3, aes(Age, as.integer(CD) - 1), 
        inherit.aes = FALSE, groupOnX = FALSE, width = 0.05, alpha = 0.7) +
    labs(y = "CD prevalence", x = "Age (years)")
```

We can see that the spline captures the higher prevalance at intermediate ages
and the possible uptick at very old ages.

Now, add the age spline to the multivariate model:
```{r}
b3 <- brm(data = tb3, family = bernoulli,
    CD ~ Ab_status + GI_status + Sex + s(Age) + CH,
    prior = base_prior,
    control = list(adapt_delta = 0.98)
)
```

Check the posteriors and chains:
```{r, fig.asp = 2.2}
plot(b3, N = 10)
```

Compare the CH coefficients of the three models:
```{r}
b1 %>% fixef(probs = c(0.05, 0.1, 0.9, 0.95), pars = c("CH1", "CH2"))
b2 %>% fixef(probs = c(0.05, 0.1, 0.9, 0.95), pars = c("CH1", "CH2"))
b3 %>% fixef(probs = c(0.05, 0.1, 0.9, 0.95), pars = c("CH1", "CH2"))
```

```{r}
list(b1, b2, b3) %>%
    map_dfr(gather_draws, `b_CH.`, regex = TRUE, .id = "model") %>%
    ggplot(aes(y = fct_rev(model), x = .value)) +
    geom_vline(xintercept = 0, color = "grey") +
    geom_halfeyeh(.width = c(0.66, 0.95)) +
    facet_wrap(~.variable, scales = "free_x")
```

The CH coefficients are consistent in the three models. The estimation error
increases and the mean estimate of the CH2 effect lessens somewhat for the
model with age, but greater shrinkage is expected from the smaller sample size.

Visualize the predictions of the full model in the absence of a GI condition,
```{r}
f3 <- crossing(
    Age = seq(0, 16, by = 0.1),
    Ab_status = c("Non-treated", "Treated"),
    Sex = c("F", "M"),
    GI_status = c("Healthy"),
    CH = factor(c(0, 1, 2)),
    ) %>%
    bind_cols(., fitted(b3, ., probs = c(0.1, 0.9)) %>% as_tibble)
f3 %>%
    ggplot(aes(Age, Estimate, color = CH)) +
    geom_ribbon(aes(ymin = Q10, ymax = Q90, fill = CH), 
        linetype = 0, alpha = 0.2) +
    geom_line() +
    facet_grid(Sex ~ Ab_status) +
    geom_quasirandom(data = tb3, aes(Age, as.integer(CD) - 1, color = CH), 
        inherit.aes = FALSE, groupOnX = FALSE, width = 0.05, alpha = 0.7) +
    ggthemes::scale_color_colorblind() +
    ggthemes::scale_fill_colorblind() +
    labs(y = "CD prevalence", x = "Age (years)") +
    theme(panel.spacing.y = unit(5, "mm"))
```

## Session info

```{r}
sessionInfo()
```
