---
title: "Separate NMDS plots for canines and equines"
author: "Mike"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
---


## Setup

```{r setup, include=FALSE}
knitr::opts_knit$set(progress = TRUE, verbose = TRUE)
# Global chunk options
knitr::opts_chunk$set(
    # cache = TRUE, autodep = TRUE,
    include = TRUE, echo = TRUE,
    warning = TRUE, message = FALSE, 
    fig.width = 6, fig.asp = 3/4
)
```

```{r load_packages}
library(phyloseq)
library(tidyverse)
library(here)
library(cowplot)
# Custom functions and plot setup
source(here("analysis", "functions-and-ggplot-setup.R"))
```

Load the microbiome profiles, excluding the single Alpaca and Avian sample.
```{r, load_data}
ps <- readRDS(here("results", "ps.Rds")) %>%
    subset_samples(Host_species %in% c("Canine", "Feline", "Equine", 
            "Ovine")) %>% 
    filter_taxa(function (x) sum(x) > 0, prune = TRUE)
ps
```
We consider a sample as CD positive if CD was detected either the lab assay or
in the microbiome sequencing profiles.
```{r}
sample_data(ps)$CD <- sample_data(ps)$CD_either
sam <- sample_data(ps) %>% as_tibble
glimpse(sam)
```

Convert profiles to proportions
```{r}
ps.prop <- ps %>%
    transform_sample_counts(function (x) x / sum(x))
```

Get separate phyloseq objects for canines and equines
```{r}
ps.canine <- ps.prop %>%
    subset_samples(Host_species == "Canine") %>%
    filter_taxa(function (x) sum(x) > 0, prune = TRUE)
ps.equine <- ps.prop %>%
    subset_samples(Host_species == "Equine") %>%
    filter_taxa(function (x) sum(x) > 0, prune = TRUE)
```

## NMDS plots

Do NMDS separately for canines and for equines
```{r}
set.seed(1234)
tb <- list(Canine = ps.canine, Equine = ps.equine) %>%
  enframe("host_species", "ps") %>%
  mutate(ord = map(ps, ordinate, method = "NMDS", distance = "bray", 
      trymax = 50))
```

Create plots
```{r}
tb <- tb %>%
  mutate(
    plot = map2(ps, ord, 
      ~ plot_ordination(.x, .y, color = "Host_species", 
        shape = "CD", type="samples") +
        scale_color_manual(values = colors.host_species, guide = NULL) +
        scale_shape_manual(values = shape.cd)
      ),
    plot = map2(plot, host_species, 
      ~ .x +
        labs(shape = italicize("C. difficile"), title = .y)
    )
  )
```

```{r, fig.dim = c(5,6) * 5/3}
plot_grid(plotlist = tb$plot, ncol = 1)
```


```{r}
ggsave(here("figures", "nmds-canine-equine-separated.pdf"), 
    width = 5, height = 6, units = "in", scale = 5/3,
    useDingbats = FALSE)
```


## Session info

```{r}
sessionInfo()
```
